SN-\sncode: A Gravitationally-Lensed Type Ia Supernova Candidate at $z=1.95$

### 2020-06-17 Paper/unused/main_backup.tex
Updates from Overleaf

\title{Three magnified views of a supernova explosion 10 billion years ago, and a fourth that will reveal the nature of cosmological dark energy}

@@

\begin{abstract}
% https://www.nature.com/documents/nature-summary-paragraph.pdf
%One or two sentences providing a basic introduction to the field, comprehensible to a scientist in any discipline
The expansion history of the Universe is one of the most fundamental observations leading to the hot big bang model for the origin of the Universe\footnote{Penzias \& Wilson 1965, ApJL, 142, 419}. 
%Two to three sentences of more detailed background, comprehensible to scientists in related disciplines.
The observed brightness of distant Type Ia supernova (SN) explosions has played a key role in mapping the expansion history, and led to the discovery of dark energy that now appears to be driving an accelerating cosmic expansion rate\footnote{Riess et al. 1998, AJ, 116, 1009; Perlmutter et al. 1999, ApJ, 517, 565; Riess et al. 2007, ApJ, 659, 98}.
Determining the nature of dark energy and how it may evolve over time is a primary goal for current and planned large-scale cosmology experiments\footnote{Amendola et al. 2013, Living Reviews in Relativity, Volume 16, Issue 1, article id. 6, 270 pp.; Ivezic et al. 2019, ApJ, 873, 111}.
%One sentence clearly stating the general problem being addressed by this particular study
One promising method for determining the dark energy equation of state and its evolution  is through time delays of gravitational lens systems, which for a given lens potential provide a measurement of the ratio of the angular diameter distances to the foreground lens and background source\footnote{Coe \& Moustakas, 2009, Apj 706, 45; Suyu 2010, ApJ, 711, 201}.
%One sentence summarising the main result (with the words “here we show” or their equivalent).
Here we report the discovery of a strongly-lensed Type Ia SN explosion that will provide a time delay precision of $<1\%$.  The SN is in an evolved galaxy at $z=1.95$ gravitationally lensed by a massive foreground galaxy cluster.
%Two or three sentences explaining what the main result reveals in direct comparison to what was thought to be the case previously, or how the main result adds to previous knowledge.
With three lensed images of the supernova detected in existing observations from the \textit{Hubble Space Telescope} with time delays of order 50 days, this is the most distant multiply-imaged supernova yet discovered.  A fourth image in the component closest to the cluster core is predicted to appear in the year 2035$\pm$2.  Observing the lightcurve of that future image will provide a time delay precision of $\approx 7$ days over an unprecedented baseline of nearly 20 years.  This single source will therefore provide a similar precision and complementary cosmological constraints to an ensemble of dozens of time delay measurements expected at somewhat lower redshifts and shorter timescales from the planned ground-based LSST and WFIRST space telescope missions.  If we suppose that cosmological parameters such as the local expansion rate ($H_0$) and dominant energy densities $\Omega_m$ and $\Omega_\Lambda$ will be well established in 2035, the combined time delay measurements will constrain the dark energy equation of state ($w_0$) to a precision of 20\%.
%One or two sentences to put the results into a more general context. 
% Considering this observation in the context of
% Two or three sentences to provide a broader perspective, readily comprehensible to a scientist in any discipline, may be included in the first paragraph if the editor considers that the accessibility of the paper is significantly enhanced by their inclusion.

@@

% \section*{Introduction}

>>>>

\title{96 ch. Three magnified views of a supernova explosion 10 billion years ago, and a fourth that will reveal the nature of cosmological dark energy}

@@

\begin{abstract}
% https://www.nature.com/documents/nature-summary-paragraph.pdf
%One or two sentences providing a basic introduction to the field, comprehensible to a scientist in any discipline
125W: The expansion history of the Universe is one of the most fundamental observations leading to the hot big bang model for the origin of the Universe\footnote{Penzias \& Wilson 1965, ApJL, 142, 419}. 
%Two to three sentences of more detailed background, comprehensible to scientists in related disciplines.
The observed brightness of distant Type Ia supernova (SN) explosions has played a key role in mapping the expansion history, and led to the discovery of dark energy that now appears to be driving an accelerating cosmic expansion rate\footnote{Riess et al. 1998, AJ, 116, 1009; Perlmutter et al. 1999, ApJ, 517, 565; Riess et al. 2007, ApJ, 659, 98}.
Determining the nature of dark energy and how it may evolve over time is a primary goal for current and planned large-scale cosmology experiments\footnote{Amendola et al. 2013, Living Reviews in Relativity, Volume 16, Issue 1, article id. 6, 270 pp.; Ivezic et al. 2019, ApJ, 873, 111}.
%One sentence clearly stating the general problem being addressed by this particular study
One promising method for determining the dark energy equation of state and its evolution  is through time delays of gravitational lens systems, which for a given lens potential provide a measurement of the ratio of the angular diameter distances to the foreground lens and background source\footnote{Coe \& Moustakas, 2009, Apj 706, 45; Suyu 2010, ApJ, 711, 201}.
%One sentence summarising the main result (with the words “here we show” or their equivalent).
Here we report the discovery of a strongly-lensed Type Ia SN explosion that will provide a time delay precision of $<1\%$.  The SN is in an evolved galaxy at $z=1.95$ gravitationally lensed by a massive foreground galaxy cluster.
%Two or three sentences explaining what the main result reveals in direct comparison to what was thought to be the case previously, or how the main result adds to previous knowledge.
With three lensed images of the supernova detected in existing observations from the \textit{Hubble Space Telescope} with time delays of order 50 days, this is the most distant multiply-imaged supernova yet discovered.  A fourth image in the component closest to the cluster core is predicted to appear in the year 2035$\pm$2.  Observing the lightcurve of that future image will provide a time delay precision of $\approx 7$ days over an unprecedented baseline of nearly 20 years.  This single source will therefore provide a similar precision and complementary cosmological constraints to an ensemble of dozens of time delay measurements expected at somewhat lower redshifts and shorter timescales from the planned ground-based LSST and WFIRST space telescope missions.  If we suppose that cosmological parameters such as the local expansion rate ($H_0$) and dominant energy densities $\Omega_m$ and $\Omega_\Lambda$ will be well established in 2035, the combined time delay measurements will constrain the dark energy equation of state ($w_0$) to a precision of 20\%.
%One or two sentences to put the results into a more general context. 
% Considering this observation in the context of
% Two or three sentences to provide a broader perspective, readily comprehensible to a scientist in any discipline, may be included in the first paragraph if the editor considers that the accessibility of the paper is significantly enhanced by their inclusion.

@@

\section*{10 paragraphs - (total 2500 words including references, notes and captions}
SN1a play a special role in cosmology, they are standard candles -> led to discovery of DE. High redshift SN-1a provide particularly strong constraints.
\newline
\newline
Multiple lensed transient or varying sources can be used to constrain cosmology, through timedelay measurements.  Many quasars lensed by galaxies known, provide some of the most accurate measurements of H0.  Possibly in dissagremment with CMB 
\newline
\newline
Refsdal proposed multiple lensed SN as means of constraining cosmology. A few have been observed (list), most famously the Refsdal SN, a Type II at z=1.5 lensed by a cluster, missed one image, observed 2 and predicted a fourth that was later observed
\newline
\newline
Requiem is a hubble project that target rare examples of quiescent galaxies, that have been magnified by gravitational lensing. These are typically too faint and compact to be resolved even with Hubble, but due to lensing are magnified. This provides a unique opportunity to  understand the still mysterious quenching mechnism. 
The brightest and most spectacular target of REQUIEM, first discovered by Newman in 2016 is RGM0138, is a z=2.X, logM=XX quiescent galaxy quadruply lensed by a cluster at z=0.X. The galaxy has very low SFR=XX and based on stellar ages stopped forming stars ~X.X Gyr earlier. This is a proto typical early type galaxy progenitor.
\newline
\newline
During analysis of the REQUIEM observations obtained  July 2019 we discovered a new pointsource close to three of the four images, which was not there in the original Hubble images taken in July 2016 (PI Newman), when the system was first discovered (Fig 1). 
\newline
\newline

The SN has different colors and brightness in the three observed images. This is expected as SN evolve in brightness and color with time, and the three images represent different times on the supernova light curve, due to the different time delays of the different images. Fig 2 shows the colors and magnitudes expected for different types of SN lightcurves. The observations classify the explosion as a Type Ia, when the magnifications from the lensing model is taken into account. This is consistent with the host galaxy being an evolved early type (void of O and B stars)
\newline
\newline
The lensing model is derived using the lenstool software, constrained by XX spectroscopically confirmed lensed images, and scaling relation for the cluster galaxies, as well as local pertubers .... (johan)
\newline
\newline
In addition to the relative magnifications of the individual images, the lensing model predicts relative timdelays between the images. These are are XXX in a greement with the observed colors evolution in Fig 2 providing confidence in the accuracy of the lensing model. The lensing model predicts that the SN should appear in the fourth image in 20XX+/-X., with a magnification of X. A fifth image will also appear in 20XX+/- but this will be demagnified and likely not observable.
\newline
\newline
This is the second highest redshift SNIa, it is the first one that where we observe it from the appearance of the first image to the last (Refsdal had an image that appeared Xyears before it was first dicovered), and it it the one with the longest (predicted) timedelay.  Add here other properties that make the system unique.
\newline
\newline
Due to the long timedelay baseline, the sensitivity to cosmology is exceptional, and the precision with which cosmological parameters can be constrained is limited only by the precision of the lensing model. Assuming that H0 is determined from other methods when the fourth image appears in 20XX Fig 3 shows that this single system will have provide comparable (and complementary) constraints as those expectede from tens of lensed supernova in planned LSST and WFIRST cosmological surveys. At that time JWST will likely be in the past but we will have E-ELT, and XXX.. which will allow for cool science such as...

### 2020-06-17 Paper/unused/main_old.tex
Updates from Overleaf

\input{Paper/preamble}

\title{96 ch. Three magnified views of a supernova explosion 10 billion years ago, and a fourth that will reveal the nature of cosmological dark energy}

\author[1,*]{Gabriel Brammer}
\author[2]{Steven Rodney}
\author[2]{Justin Pierel}
\author[2]{Kyle O'Connor}
\author[3]{Johan Richard}
\author[1]{Sune Toft}
\author[4]{Mohammad Akhshik}
\author[5]{Katherine Whitaker}
\affil[1]{Cosmic Dawn Center, Niels Bohr Institute, K{\o}benhavns Universitet}
\affil[2]{University of South Carolina}
\affil[3]{Universit\'e Claude Bernard Lyon}
\affil[4]{University of Connecticut}
\affil[5]{University of Massachusetts, Amherst}

@@

@@

\end{abstract}
\begin{document}

@@

\section*{10 paragraphs - (total 2500 words including references, notes and captions}
SN1a play a special role in cosmology, they are standard candles -> led to discovery of DE. High redshift SN-1a provide particularly strong constraints.
\newline
\newline
Multiple lensed transient or varying sources can be used to constrain cosmology, through timedelay measurements.  Many quasars lensed by galaxies known, provide some of the most accurate measurements of H0.  Possibly in dissagremment with CMB 
\newline
\newline
Refsdal proposed multiple lensed SN as means of constraining cosmology. A few have been observed (list), most famously the Refsdal SN, a Type II at z=1.5 lensed by a cluster, missed one image, observed 2 and predicted a fourth that was later observed
\newline
\newline
Requiem is a hubble project that target rare examples of quiescent galaxies, that have been magnified by gravitational lensing. These are typically too faint and compact to be resolved even with Hubble, but due to lensing are magnified. This provides a unique opportunity to  understand the still mysterious quenching mechnism. 
The brightest and most spectacular target of REQUIEM, first discovered by Newman in 2016 is RGM0138, is a z=2.X, logM=XX quiescent galaxy quadruply lensed by a cluster at z=0.X. The galaxy has very low SFR=XX and based on stellar ages stopped forming stars ~X.X Gyr earlier. This is a proto typical early type galaxy progenitor.
\newline
\newline
During analysis of the REQUIEM observations obtained  July 2019 we discovered a new pointsource close to three of the four images, which was not there in the original Hubble images taken in July 2016 (PI Newman), when the system was first discovered (Fig 1). 
\newline
\newline

The SN has different colors and brightness in the three observed images. This is expected as SN evolve in brightness and color with time, and the three images represent different times on the supernova light curve, due to the different time delays of the different images. Fig 2 shows the colors and magnitudes expected for different types of SN lightcurves. The observations classify the explosion as a Type Ia, when the magnifications from the lensing model is taken into account. This is consistent with the host galaxy being an evolved early type (void of O and B stars)
\newline
\newline
The lensing model is derived using the lenstool software, constrained by XX spectroscopically confirmed lensed images, and scaling relation for the cluster galaxies, as well as local pertubers .... (johan)
\newline
\newline
In addition to the relative magnifications of the individual images, the lensing model predicts relative timdelays between the images. These are are XXX in a greement with the observed colors evolution in Fig 2 providing confidence in the accuracy of the lensing model. The lensing model predicts that the SN should appear in the fourth image in 20XX+/-X., with a magnification of X. A fifth image will also appear in 20XX+/- but this will be demagnified and likely not observable.
\newline
\newline
This is the second highest redshift SNIa, it is the first one that where we observe it from the appearance of the first image to the last (Refsdal had an image that appeared Xyears before it was first dicovered), and it it the one with the longest (predicted) timedelay.  Add here other properties that make the system unique.
\newline
\newline
Due to the long timedelay baseline, the sensitivity to cosmology is exceptional, and the precision with which cosmological parameters can be constrained is limited only by the precision of the lensing model. Assuming that H0 is determined from other methods when the fourth image appears in 20XX Fig 3 shows that this single system will have provide comparable (and complementary) constraints as those expectede from tens of lensed supernova in planned LSST and WFIRST cosmological surveys. At that time JWST will likely be in the past but we will have E-ELT, and XXX.. which will allow for cool science such as...

@@

% Figures and tables can be referenced in LaTeX using the ref command, e.g. Figure \ref{fig:stream} and Table \ref{tab:example}.

\end{document}

>>>>

\usepackage{xspace}
\usepackage{multirow}
\usepackage[numbers]{natbib}
\usepackage[colorinlistoftodos,textwidth=1.4in,textsize=small]{todonotes}

% 96 characters including spaces
\title{(Requiem for a supernova 10 billion years ago, which in 2030 will reveal the nature of dark energy}

\author[1,2,*]{Gabriel Brammer}
\author[3]{Steven Rodney}
\author[4]{Justin Pierel}
\author[4]{Kyle O'Connor}
\author[4]{Johan Richard}
\author[1,2]{Sune Toft}
\author[5]{Mohammad Akhshik}
\author[6,1]{Katherine Whitaker}
\affil[1]{Cosmic Dawn Center (DAWN)}
\affil[2]{ Niels Bohr Institute, University of Copenhagen, Denmark}
\affil[3]{University of South Carolina}
\affil[5]{Universit\'e Claude Bernard Lyon}
\affil[6]{University of Connecticut}
\affil[6]{University of Massachusetts, Amherst}

@@

%https://www.sciencemag.org/authors/instructions-preparing-initial-manuscript

%a sentence giving a broad introduction to the field comprehensible to the general reader,

%a sentence of more detailed background specific to your study.

%an explanation of the OBJECTIVES/METHODS, and then the RESULTS

%The final sentence should outline the main CONCLUSIONS of the study, in terms that will be comprehensible to all our readers

@@

\end{abstract}

\begin{document}
\listoftodos

@@

\section*{10 paragraphs - (total 2500 words including references, notes and captions}
\todo[inline]{JP: Can edit intro} SN1a play a special role in cosmology, they are standard candles -> led to discovery of DE. High redshift SN-1a provide particularly strong constraints.
\newline
\newline
Multiple lensed transient or varying sources can be used to constrain cosmology, through timedelay measurements.  Many quasars lensed by galaxies known, provide some of the most accurate measurements of H0.  Possibly in dissagremment with CMB 
\newline
\newline
Refsdal proposed multiple lensed SN as means of constraining cosmology. A few have been observed (list), most famously the Refsdal SN, a Type II at z=1.5 lensed by a cluster, missed one image, observed 2 and predicted a fourth that was later observed
\newline
\newline
Requiem is a hubble project that target rare examples of quiescent galaxies, that have been magnified by gravitational lensing. These are typically too faint and compact to be resolved even with Hubble, but due to lensing are magnified. This provides a unique opportunity to  understand the still mysterious quenching mechnism. 
The brightest and most spectacular target of REQUIEM, first discovered by Newman in 2016 is RGM0138, is a z=2.X, logM=XX quiescent galaxy quadruply lensed by a cluster at z=0.X. The galaxy has very low SFR=XX and based on stellar ages stopped forming stars ~X.X Gyr earlier. This is a proto typical early type galaxy progenitor.
\newline
\newline
During analysis of the REQUIEM observations obtained  July 2019 we discovered a new pointsource close to three of the four images, which was not there in the original Hubble images taken in July 2016 (PI Newman), when the system was first discovered (Fig 1). 
\newline
\newline

\todo[inline]{JP: I'll fill this in from the old text, then SR and/or KO work on this classification section} The SN has different colors and brightness in the three observed images. This is expected as SN evolve in brightness and color with time, and the three images represent different times on the supernova light curve, due to the different time delays of the different images. Fig 2 shows the colors and magnitudes expected for different types of SN lightcurves. The observations classify the explosion as a Type Ia, when the magnifications from the lensing model is taken into account. This is consistent with the host galaxy being an evolved early type (void of O and B stars). Using the absolute magnitude, $M_K$, and color, $B-K$, of the host, there is about 75 percent probability that the object is of type Ia (citation to https://arxiv.org/pdf/1309.2630.pdf).
\newline
\newline
\todo[inline]{JP: Need Johan to fill in this section, or give details for us to fill it in} The lensing model is derived using the lenstool software, constrained by XX spectroscopically confirmed lensed images, and scaling relation for the cluster galaxies, as well as local pertubers .... (johan)
\newline
\newline
\todo[inline]{JP: I'll fill this in + Johan lens model stuff, this will include time delay/magnification constraints from SNTD} In addition to the relative magnifications of the individual images, the lensing model predicts relative timdelays between the images. These are are XXX in a greement with the observed colors evolution in Fig 2 providing confidence in the accuracy of the lensing model. The lensing model predicts that the SN should appear in the fourth image in 20XX+/-X., with a magnification of X. A fifth image will also appear in 20XX+/- but this will be demagnified and likely not observable.
\newline
\newline
This is the second highest redshift SNIa, it is the first one that where we observe it from the appearance of the first image to the last (Refsdal had an image that appeared Xyears before it was first dicovered), and it it the one with the longest (predicted) timedelay.  Add here other properties that make the system unique.
\newline
\newline
\todo[inline]{JP: Can edit time delay/cosmo section} Due to the long timedelay baseline, the sensitivity to cosmology is exceptional, and the precision with which cosmological parameters can be constrained is limited only by the precision of the lensing model. Assuming that H0 is determined from other methods when the fourth image appears in 20XX Fig 3 shows that this single system will have provide comparable (and complementary) constraints as those expectede from tens of lensed supernova in planned LSST and WFIRST cosmological surveys. At that time JWST will likely be in the past but we will have E-ELT, and XXX.. which will allow for cool science such as...

@@

% Figures and tables can be referenced in LaTeX using the ref command, e.g. Figure \ref{fig:stream} and Table \ref{tab:example}.
\pagebreak
\citep{jha_improved_2007}
\bibliographystyle{plainnat}
\bibliography{Paper/references}
\end{document}

### 2020-06-17 Paper/unused/ms.tex
Updates from Overleaf

\correspondingauthor{Gabriel Brammer}
\email{gabriel.brammer@nbi.ku.dk}

@@

To evaluate the age of each transient image (and therefore constrain the lensing time delay between images), it is valuable to have a firm determination of the transient's class.  The range of possible transients is broad, encompassing normal SNe, superluminous SNe \citep{galyam_luminous_2019}, fast transients \citep{berger_fast_2013, drout_rapidly_2014}, and even gravitational microlensing events \citep{rodney_two_2018,kelly_extreme_2018}.  Identifying the transient type makes it possible to use the well-developed library of spectrophotometric SN models \citep[e.g.][]{kessler_models_2019} for the time delay measurement. 
 The first-order distinction that is most important is between a Type Ia SN--understood to be the explosion of a white dwarf star in a binary system--and core collapse SN (CCSN)--the end-point of a star with mass $\gtrsim 10$M$_{\odot}$.  Given the limited data available for this transient, we will not attempt a more fine-grained classification, though in principle that could be achieved with spectroscopy and multi-band photometry upon arrival of the final image.

>>>>

\correspondingauthor{TBD}
\email{tbd@aol.com}

@@

To evaluate the age of each transient image (and therefore constrain the lensing time delay between images), it is valuable to have a firm determination of the transient's class.  The range of possible transients is broad, encompassing normal SNe, superluminous SNe \citep{galyam_luminous_2019}, fast transients \citep{berger_fast_2013, drout_rapidly_2014}, and even gravitational microlensing events \citep{rodney_two_2018,kelly_extreme_2018}.  Identifying the transient type makes it possible to use the well-developed library of spectrophotometric SN models \citep[e.g.][]{kessler_models_2019} for the time delay measurement.
